\chapter{The finite element method}
\label{sec:fem}

This chapter discusses the well known and widely used method for solving partial differential equations, the finite element method (FEM).
To avoid building and adding another library to the already vast ecosystem of available FEM software, we use the multi-physics FEM library, \oomph.
From the website, \oomph is:
\begin{quote}
	an object-oriented, open-source finite-element library for the simulation of multi-physics problems. \cite{oomph}
\end{quote}
There are several high level features of any general finite element procedure that will be addressed below.
Discussing these features in generality allows for specificity later to any problem.
However, we will also be explicit with description about the problem referenced earlier.
In addition to this, an overview of specific implementation details and features available in \oomph will also be covered.





% -----------------------------------------------------------

\section{Preliminaries}

\iffalse

.Problem
.Classical vs weak solution
Test functions
Sobolev space (where test functions live)
Basis/shape functions
Galerkin method -> test functions are basis functions

\fi

We begin this chapter by discussing the necessary mathematical background for the finite element method.

The classical or strong solution of a problem is a function $u_N(r,z)$ satisfying the PDE
\begin{align}
	\nabla^2 u_N + \left[ (1+i\alpha)k^2-\frac{N^2}{r^2}\right]u_N = 0,
\end{align}
at every point in the domain $\Omega$ and satisfying the boundary conditions
\begin{align}
	u_N(r,z) \vert_{\partial \Omega_D} = g_D \\
	\frac{\partial u_N(r,z)}{\partial \vec{n}} \vert_{\partial \Omega_N} = g_N
\end{align}
at every point on the respective portion of the boundary $\partial\Omega$.

The weak solution of a problem is a relaxation of the conditions for the existence of a solution to the problem.
A weak solution $u_N^{\mathrm{weak}}(r,z)$ satisfies the essential (Dirichlet) boundary condition and that
\begin{align}
	\int_\Omega \left( \nabla^2 u_N + \left[ (1+i\alpha)k^2-\frac{N^2}{r^2}\right]u_N \right) v = 0
\end{align}
for all test functions $v$ that are 0 on the essential boundary.

The test functions are used to weight the integral of the residual equation and are members of the function space
\begin{align}
	H^1_0(\Omega) = \left\{ v | v=0 \mathrm{on} \partial\Omega_D, \, v \in L^2(\Omega), \, \nabla v \in (L^2(\Omega))^2 \right\}.
\end{align}
The solution $u_N^{\mathrm{weak}}(r,z)$ must satisfy the Dirichlet boundary condition, and so is a member of the space
\begin{align}
	H^1(\Omega) = \left\{ v | v \in L^2(\Omega), \, \nabla v \in (L^2(\Omega))^2 \right\}
\end{align}
The function spaces $H^i(\Omega)$ are known as Sobolev spaces and are complete normed vector spaces.


% For the strong form above, we require at least that $u_N(r,z) \in C^2(\Omega)$.
% A sufficiently smooth solution satisfying the weak form will also satisfies the stong form.
% Therefore the two statements of the problem are equivalent, given the smoothness requirements of the classical formulation.


The solution on the mesh is represented using shape functions that form a basis for the domain.

For a two-dimensional domain, the shape functions are bilinear and 
In particular, for the global node $j$ in the mesh, the basis function $\phi_j(r,z)$ is 1 at node $j$ and 0 everywhere else.


Local vs global coordinates




% -----------------------------------------------------------

\section{Nodes, elements, and meshes}

The main components of the finite element method are the nodes, the elements, and the mesh.
A mesh is a collection of elements; an element is a collection of nodes.

Each node contains data about its location within the mesh, its local coordinate within the element, as well as the value of the solution at the node.

The elements are the 

The number of nodes of an element determines its order.
If For example, an element with

How do we get from global to local coordinates?
We need the mapping
And with that we need the Jacobian of the mapping
(we have two change of coordinates: Cartesian -> Cylindrical -> Local
And so we have two Jacobian mappings)

On the elements we have defined basis functions.
These represent the function 

Together with the elements, other fundamental pieces of the FEM are the nodes and the mesh.
Every element consists of multiple nodal points, each of which holds data required for solving the problem.
For example, nodes contain the global coordinate of their position within the mesh



The FEM lends itself well to object-oriented programming (OOP) design principles.
This is due to the discrete entities involved and the relationships between them.
Such a design strategy involves the definition of classes, and their instantiation as objects.
Furthermore, classes may be extended using inheritance, allowing for highly abstracted and generic classes, and on the other end, highly specific classes.
This is an advantage for both the developers and the users of \oomph.
Developers are able to work independently of each other, knowing only how objects should interact.
Users are able to extend previous work to suit their own needs.






% -----------------------------------------------------------

\section{Weak formulation}

We will now derive the weak form of the CSLP equation, \eqref{eqn:cslp}.
Our first step is to multiply the equation by a test function $v\in H^1_0(\Omega)$ and integrate over the domain $\Omega$.
\begin{align}
	-\int_\Omega \left(\nabla^2 u_N + \left[ (1+i\alpha)k^2-\frac{N^2}{r^2}\right]u_N \right) v = 0.
\end{align}
Using the property of the gradient operator,
\begin{align}
	\nabla \cdot ( v \nabla u) = v \nabla^2 u + \nabla u \cdot \nabla v,
\end{align}
we arrive at
\begin{align}
	\int_\Omega \nabla u_N \cdot \nabla v 
  - \int_\Omega \nabla \cdot (v \nabla u ) 
  - \int_\Omega \left[(1+i\alpha)k^2 - \frac{N^2}{r^2}\right] u_N v = 0.
\end{align}

Using the Gauss-Divergence theorem, we may write the second term as an integral over the boundary of the domain $\partial \Omega$,
\begin{align}
	\int_\Omega \nabla \cdot (v \nabla u ) = \int_{\partial\Omega} v \frac{\partial u_N}{\partial \vec{n}},
\end{align}
where $\vec{n}$ is the outward facing normal vector from the boundary.
The boundary integral may be further split up into parts sorted by their respective boundary conditions.
If the portion of the boundary whose solution value is known is represented by $\partial \Omega_D$ and the remaining boundary is represented as $\partial \Omega_N$, then
\begin{subequations}
\begin{align}
	\partial \Omega_D \cup \partial \Omega_N &= \partial \Omega. \\
	\partial \Omega_D \cap \partial \Omega_N &= \emptyset.
\end{align}
\end{subequations}
Given these exclusivity properties on the boundary, we may write the term for the boundary integral as
\begin{align}
	\int_{\partial\Omega} = \int_{\partial\Omega_D} + \int_{\partial\Omega_N}.
\end{align}
Since by definition, the test functions are exactly 0 on the Dirichlet portion of the boundary, it is the case that
\begin{align}
	\int_{\partial\Omega_D} v \frac{\partial u_N}{\partial \vec{n}} = 0.
\end{align}
We will also consider a boundary with only Dirichlet conditions, so that the Neumann part also vanishes.

Combining the results, we arrive at the weak form: find $u_N(r,z) \in H^1(\Omega)$ such that for all $v \in H^1_0(\Omega)$,
\begin{align}
% 	\int_\Omega \nabla u_N \cdot \nabla v 
%   - \int_{\partial\Omega_N} \nabla \cdot (v \nabla u_N ) 
%   - (1+i\alpha)k^2 \int_\Omega u_N v 
%   - N^2 \int_\Omega \frac{1}{r^2} u_N v = 0. \label{eqn:weakform}
	\int_\Omega \left[
		\nabla u_N \cdot \nabla v 
	  - \left((1+i\alpha)k^2 - \frac{N^2}{r^2}\right) u_N v 
	\right] = 0. \label{eqn:weakform}
\end{align}


One further point regarding the implementation of this within \oomph.
Complex numbers are not handled natively within \oomph.
Instead, a two-dimensional vector whose first and second components store the real and imaginary parts of a complex number, respectively.
Therefore to deal with this, we split the solution $u_N$ into its real and imaginary components.
If $u_r$ is the real part of the solution and $u_c$ is the imaginary part, then
\begin{align}
	u_N = u_r + i u_c.
\end{align}
Substituting this into \eqref{eqn:weakform} and separating real and imaginary parts will acheive this.







% ----------------------------------------------------------

\section{Numerical integration}
\label{sec:integration}

The integrals involved in computing the solution are often too complicated evaluate analytically.
Instead, we use a numerical integration scheme to perform the integration.
Several methods exist to complete this task, but within \oomph, it is done using Gauss quadrature.

\cite{oomph}
Fix the dimension to be 2, so that we can talk specifics...
The Gauss rules are defined by following three quantities:
\begin{enumerate}
	\item the number of integration points, $N_\mathrm{int}$,
	\item the position of the integration points within an element, $S_i, \, i=1,2,\ldots,N_\mathrm{int}$,
	\item the integration weights, $W_i, \, i=1,2,\ldots,N_\mathrm{int}$.
\end{enumerate} 
Given these values, we can approximate an integral by summation of the integrand evaluated at the integration points multiplied by the corresponding integration weight.

We also must multiply by the Jacobian of the mapping from Cartesian to cylindrical coordinates.
Recall equation \eqref{eqn:cylindrical_mapping}, which states that the Jacobian of this transformation is $r$.
In addition, we have the transformation from global to local coordinates.
This mapping is given by some other function

Denote by $J$ the Jacobian of the mapping from global to local coordinates, then for an arbitrary element $E$, 
\begin{align}
	\iint_E f(r,z) r \,\d r \,\d z = \int_{-1}^1 \int_{-1}^1  f(s_1, s_2) J \, \d s_1 \, \d s_2 = \sum_{i=1}^{N_\mathrm{int}} f(S_{i,1},S_{i,2}) 
\end{align}






% ---------------------------------

\section{Refinement}

\iffalse
How are the meshes created?
What is the process of refinement?
	Adaptive vs uniform
Hanging nodes..
p-refinement: order of the elements
h-refinement: density/size of the mesh
Use this to generate the grids for Multigrid
\fi

There are several variations of refinement that can improve the accuracy of the finite element solution.
In addition to gains in solution accuracy, refinement is the method that we will generate the grid hierarchy for multigrid.
This is done by performing multiple refinements to achieve the finest level of refinement, then unrefining to obtain each coarse level.
More discussion of the grids used in multigrid is in chapter \ref{sec:mg}; here we will discuss the specific use of refinement within the FEM.
The largest scale of the solution can be at most at the scale of the mesh, so often refinement is necessary to resolve the exact features of the solution.
Each method of refinement comes with costs and benefits, as well as different complexities in implementation.

% Uniform refinement
The simplest case is uniform h-refinement.
In h-refinement, elements are subdivided into multiple new elements, reducing the element width and increasing the resolution of the mesh.
In uniform refinement, the entire mesh is refined by subdividing every element.
This increases the storage and computational cost of the method, but leads to a reduction in the solution error.
Figure \ref{fig:uniform} shows an example of a mesh consisting of four elements on the left and the result on the right after the application of uniform h-refinement.
We can see here that each element has been split into four new elements, each with element width half of the original element.

% Adaptive refinement
Beyond uniform refinement is adaptive refinement.
In certain cases where parts of the solution domain are singular or
In problems such as those involving boundary layers, for example in fluid dynamics, the solution may be highly varying in only one specific region.
By using an error estimator, it is possible to determine the areas of the mesh that require refinement.
Using uniform refinement on this problem, the grid size of the mesh will be reduced across the whole mesh, increasing computational cost.
Adaptive refinement has the benefit that computation is not wasted on computing the solution in regions where a sufficiently large mesh size is more appropriate.

However, adaptive refinement can introduce hanging nodes to the mesh.
Hanging nodes occur when a node is not shared between surrounding elements.
The problem with this type of node is that inter-element continuity of the solution is lost.
To overcome hanging nodes, we must impose constraints on the nodal values so that continuity between elements is maintained.
We will focus only on uniform refinement in this project and not discuss further adaptive refinement.

Figure \ref{fig:adaptive} shows an example of a mesh consisting of four elements on the left and the result on the right after the application of adaptive h-refinement.
On the right mesh, the element in the top right corner has been refined, while the others remain the same.
This has introduced two hanging nodes to the element, displayed here in red.

Alternatively, there is also p-refinement.
This method of refinement increases the order of the element by increasing the number of nodes.
Recall that we can find an interpolating polynomial that passes through $n+1$ distinct points with a polynomial of degree $n$.
So for example, if there are two nodes in a one-dimensional slice of an element, a linear polynomial can interpolate the solution between them.
If the solution is highly nonlinear, increasing the order of the interpolating polynomial will decrease the interpolation error.
Hence, increasing the number of nodes can lead to a reduction in the solution error.
\begin{figure}[h]
	% Uniform and adaptive refinement

	\centering
	
	\subfloat[][\label{fig:uniform} Uniform refinement of a mesh.]{\scalebox{0.4}{
		\begin{tikzpicture}[baseline,decoration={markings,mark=at position 1 with
			{\arrow[scale=5,>=stealth]{>}}}]
		\centering
		\begin{scope}
		
		% define constants
		\def \w {8}
		\def \d {4}
		
		% Draw the grids
		\draw[step=4,black,very thin] (0,0) grid (\w,\w);
		\draw[postaction={decorate}] (\w+1,\w/2) -- (\w+\d-1,\w/2);

		\draw[step=2,black,very thin] (\w+\d,0) grid (2*\w+\d,8);
		
		\end{scope}
		\end{tikzpicture}
	}}\\

	\subfloat[][\label{fig:adaptive} Adaptive refinement of a mesh; hanging nodes are displayed in red.]{\scalebox{0.4}{
		\begin{tikzpicture}[baseline,decoration={markings,mark=at position 1 with
			{\arrow[scale=5,>=stealth]{>}}}]
		\centering
		\begin{scope}
		
		% define constants
		\def \w {8}
		\def \d {4}
		
		% Draw the grids
		\draw[step=4,black,very thin] (0,0) grid (\w,\w);
		\draw[postaction={decorate}] (\w+1,\w/2) -- (\w+\d-1,\w/2);


		\draw[step=4,black,very thin] (\w+\d,0) grid (2*\w+\d,\w);
		\draw[step=2,black,very thin] (1.5*\w+\d,\w/2) grid (2*\w+\d,\w);

		\node [red,scale=2] at (1.5*\w+\d,0.75*\w) {\textbullet};
		\node [red,scale=2] at (1.75*\w+\d,0.5*\w) {\textbullet};
		
		\end{scope}
		\end{tikzpicture}
	}}

	\caption{\label{fig:refinement} An example of uniform and adaptive h-refinement on a mesh.}
\end{figure}







% ---------------------------------

\section{Computing the finite element solution}

\iffalse
Write in residual form
- don't be equal to 0, easier to write equations.

.We're given an infinite basis.
.Expand the solution in terms of the basis functions.
.Substitute this sum into the weak form.
.Represent test functions as sum of basis functions (Galerkin).
Perform truncation of the series.
Compute the weighted residual using numerical integration.
Compute the Jacobian matrix.
Solve the linear system (MG).
Obtain the finite element solution.
\fi

Given an infinite set of basis functions $\{ \phi_k \}_{k=1}^\infty$ for $H^1_0(\Omega)$, we can represent any member of this space in terms of the basis.
We can split the solution into two parts, the first satisfying the homogeneous boundary $u_h(r,z)$, and the second satisfying the Dirichlet boundary $u_p(r,z)$.
\begin{align}
	u_N(r,z) = u_p + u_h = u_p + \sum_{j=1}^\infty u_j \phi_j.
\end{align}
for some unknown coefficient values $u_j$.
Ignoring $u_p$ for the minute, if we substitute this form of the solution into equation \eqref{eqn:weakform} and simplify, we obtain an equation for the residual $r$ in terms of the unknowns $u_j$,
\begin{align}
	r(u_1,u_2,\ldots) = \sum_{j=1}^\infty u_j \int_\Omega \left[ \nabla \phi_j \cdot \nabla v - \left( (1+i\alpha)k^2 + \frac{N^2}{r^2}\right) \phi_j v \right]. \label{eqn:galerkin}
\end{align}

Since the test function $v$ is a member of the space spanned by the basis functions, we can expand the test function in terms of this basis,
\begin{align}
	v = \sum_{i=1}^\infty v_i \phi_i.
\end{align}
And because the weak form must be satisfied for all test functions $v$, it is the case that the weak form must be satisfied for all values of the coefficients $v_i$.
Substituting this form of the test function into equation \eqref{eqn:galerkin}, the following must be satisfied for $i=1,2,\ldots$
\begin{align}
	r_i(u_1,u_2,\ldots)=\sum_{j=1}^\infty u_j \int_\Omega \left[ \nabla \phi_j \cdot \nabla \phi_i - \left( (1+i\alpha)k^2 + \frac{N^2}{r^2}\right) \phi_j \phi_i \right].
\end{align}


For the purposes of implementation, a truncation must be performed on these infinite series.
This leads to an approximation of the true solution.
Suppose that we reduce our basis set to $M$ functions.
Then we have a $M\times M$ system of equations 
\begin{align}
	\sum_{j=1}^M u_j \int_\Omega \left[ \nabla \phi_j \cdot \nabla \phi_i - \left( (1+i\alpha)k^2 + \frac{N^2}{r^2}\right) \phi_j \phi_i \right],
\end{align}
for $i=1,2,\ldots,M$.



The integrals can be evaluated using quadrature rules mentioned in section \ref{sec:integration}.
The Jacobian matrix


that can now be solved to determine the coefficients $u_j$ in the solution.




Newton's method is the general purpose solver used in \oomph to find the finite element solution.
To solve the general nonlinear problem arising from a finite element discretisation, an iterative method must be used to determine the solution.
For our specific case the problem is linear, so Newton's method will converge in exactly one iteration to the exact solution in floating point arithmetic.

Once the vector of residuals $b$, and the Jacobian matrix $A$, have been computed, all that remains is to calculate the final solution by solving the linear system.
\begin{align}
	A x = b
\end{align}
The default solver within \oomph for finding the solution to this linear system is the direct solver SuperLU <<CITATION>>.
This is the point that we swap the default solver with our implementation of the preconditioned multigrid solver.
